Binary Latent Representations for Efficient Ranking

### 2017-06-17 paper/paper.tex
Mention simplicity of implementation.

To alleviate these constraints in both online and offline settings, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score and requiring a fraction of the memory. In MovieLens 100k experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take only 6\% of the memory.

@@

We use two loss functions to train the model:
\begin{itemize}
\item Bayesian personalised ranking (BPR, \citet{rendle2009bpr}), and
\item adaptive sampling maximum margin loss, following \citet{weston2011wsabie}.
\end{itemize}
For both loss functions, for any known positive user-item interaction pair $(u, i)$, we uniformly sample an implicit negative item $j$ \in \symUserInteractionSet^-$. For BPR, the loss for any such triplet is given by
\begin{equation}
1 - \sigma\left(r_{ui} - r_{uj}\right),
\end{equation}
where $\sigma$ denotes the sigmoid function.
The adaptive sampling loss is given by
\begin{equation}
\abs{1 - r_{ui} + r_{uj}}_{+}.
\end{equation}
For any $(u, i)$ pair, if the sampled negative item $j$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a total of $k$ attempts. This leads the model to perform more gradient updates in areas where its ranking performance is poorest.

@@

\subsection{Implementation}
The model is implemented in PyTorch and trained using the nVidia K40 GPUs. During training, the embedding parameters are stored as floats. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule.

The prediction code runs on the CPU and is implemented using Intel X86 AVX2 SIMD intrinsics. SIMD (Single Instruction Multiple Data) instructions allow the CPU to operate on multiple pieces of data in parallel, achieving significant speedups over the scalar version. We use explicit intrinsics rather than compiler autovectorization to ensure that neither the real-valued nor the binary prediction code is unfairly disadvantaged by the quality of compiler autovectorization.

The real-valued prediction code is implemented using 8-float wide fused multiply-add instructions (\texttt{\_mm256\_fmadd\_ps}).

In the binary version, the real-valued embedding parameters used in training are discarded, and the derived 1-bit weights are packed into 32-bit integer buffers. The XNOR dot product is implemented using 8-integer wide XOR operations (256 binary weights are processed at a time), followed by a popcount instruction to count the number of on bits in the result. We use the \texttt{libpopcnt} \citep{mula2016faster} library for the bit counting operations.

@@

The source code for the implementation is available on Github at \url{https://github.com/maciejkula/binge}.

@@

\section{Conclusion}
We show that binary user and item representations allow recommender system designers to exploit an attractive new set of speed-accuracy trade-offs. They are straightforward to estimate, and can easily be used in any existing model estimated using stochastic gradient descent.

>>>>

To alleviate these constraints in both online and offline settings, we propose the use of binary item and user representations in learning-to-rank matrix factorization models. Following the approach of \citet{rastegari2016xnor}, we show that we can estimate binary user and item representations that achieve comparable ranking accuracy to standard learning-to-rank approaches while being several times faster to score, requiring a fraction of the memory, and being only marginally more complex to estimate. In MovieLens 100k experiments, our binary representations are within 4\% of the accuracy of the real-valued model, but are over 13 times faster to score and take only 6\% of the memory.

@@

We use two loss functions to train the model:
\begin{itemize}
\item Bayesian personalised ranking (BPR, \citet{rendle2009bpr}), and
\item adaptive sampling maximum margin loss, following \citet{weston2011wsabie}.
\end{itemize}
For both loss functions, for any known positive user-item interaction pair $(u, i)$, we uniformly sample an implicit negative item $j \in \symUserInteractionSet^-$. For BPR, the loss for any such triplet is given by
\begin{equation}
1 - \sigma\left(r_{ui} - r_{uj}\right),
\end{equation}
where $\sigma$ denotes the sigmoid function.
The adaptive sampling loss is given by
\begin{equation}
\abs{1 - r_{ui} + r_{uj}}_{+}.
\end{equation}
For any $(u, i)$ pair, if the sampled negative item $j$ results in a zero loss (that is, the desired pairwise ordering is not violated), a new negative item is sampled, up to a total of $k$ attempts. This leads the model to perform more gradient updates in areas where its ranking performance is poorest.

@@

\subsection{Implementation}
The model is implemented in PyTorch and trained using the nVidia K40 GPUs. The implementation is only marginally more complex than a standard learning-to-rank model, and is accomplished by simply replacing the user-item vector dot product operation with its binary counterpart. The change results in a small increase in training time.

During training, the embedding parameters are stored and updated as single precision floating point values. Similarly, the XNOR dot product is carried out using floats in $\{1, -1\}$. The minibatch size and number of training epochs are treated as model hyperparameters. All models are trained using the Adam training rate schedule \citep{kingma2014adam}.

The prediction code runs on the CPU and is implemented in C using Intel X86 AVX2 SIMD intrinsics. SIMD (Single Instruction Multiple Data) instructions allow the CPU to operate on multiple pieces of data in parallel, achieving significant speedups over the scalar version. We use explicit intrinsics rather than compiler autovectorization to ensure that neither the real-valued nor the binary prediction code is unfairly disadvantaged by the quality of compiler autovectorization.

The real-valued prediction code is implemented using 8-float wide fused multiply-add instructions (\texttt{\_mm256\_fmadd\_ps}). In the binary version, the real-valued embedding parameters used in training are discarded, and the derived 1-bit weights are packed into 32-bit integer buffers. The XNOR dot product is implemented using 8-integer wide XOR operations (256 binary weights are processed at a time), followed by a popcount instruction to count the number of on bits in the result. We use the \texttt{libpopcnt} \citep{mula2016faster} library for the bit counting operations.

@@

The source code for both model training and prediction is available on Github at \url{https://github.com/maciejkula/binge}.

@@

\section{Conclusion and future work}
We show that binary user and item representations allow recommender system designers to exploit an attractive new set of speed-accuracy trade-offs. They are straightforward to estimate, and can easily be used in any existing model estimated using stochastic gradient descent.
